### 2019-09-15 ic-journal/response-letter.tex

  In the formalisation you only use contexts with one kind of hole $[]$. This definition can be changed to only
  use one hole as well, with a clear indication that this hole can occur multiple times.
\end{quote}

We have completely rewritten this part, now we adope free variables
and unified CCS expressions for both processes and equations.
 Thus the hole is not needed here. (It's formal defined later and used
 to formalize the univariate version of unique-solution theorems.

\item \begin{quote}
    L82: Remove the comma after $Q'$
  \end{quote}
  Done.

\item \begin{quote}
    Definition 2.2: Superscript $c$ is already used for contextual closure.
    Does $\approx^c$ stand for the contextual closure of $\approx$, or is it something else?
    Since contextual closure is not used anywhere in the paper, it is best to remove it.
    Otherwise you need different notation for rooted bisimilarity.
  \end{quote}
  We have removed the definition of ``context closure'' since it's not
  (explicitly) used in the paper.  Then $\rapprox$ is a standard
  notation for rooted bisimilarity following
  Gorrieri~\cite{Gorrieri:2015jt}, in which the superscript $c$ should be
  understood as a separate function or operator.

@@

\item \begin{quote}
    General: in the lines 293-300 the trailing deadlock 0 is given,
    while in the preliminaries you stated that it would be omitted. If
    you decide to keep the deadlock, it should also be in bold font.
  \end{quote}
  We have made them all in bold fonts.

\item \begin{quote}
    L296: I don’t understand fact (1).
  \end{quote}
We have re-phrased this part: ``(1) there exist indeed three
  transitions mentioned above; (2) there exists no other
  transition.''

\item \begin{quote}
    L299: ‘Hence ... CCS_TRANS_CONV’ is not a grammatically correct sentence; there is a verb missing.
  \end{quote}
  We have rephrased this sentence.

\item \begin{quote}
    L316: Explain what is p.
  \end{quote}
  We have explained it in the paper. In this case there is no hole in the context.

### 2019-09-15 ic-journal/conclusions.tex
Literally handled ALL review issues. DONE.

In this paper, we have highlighted a formalisation of the theory of CCS in the 
HOL4 theorem prover.
%  (for lack of space we have not discussed 
% the formalisation of some basic algebraic theory, of the basic
% properties of the expansion preorder,   and of a few
%  versions of `bisimulation up to'
% techniques). %
%such as  bisimulation up-to bisimilarity). 
The formalisation supports 4 methods for establishing (strong and weak) bisimilarity
results: 
\begin{enumerate}
\item
 constructing a bisimulation (the standard bisimulation proof
method);
\item constructing a `bisimulation up-to'; 
\item employing algebraic laws;
\item defining a system of equations or contractions
(i.e., the `unique-solution' method)
\end{enumerate}

The formalisation has actually focused on the theory of
unique solution of equations and contractions. It    
 has also allowed us to further develop the theory,
notably the basic properties of rooted contraction, and the unique
solution theorem for it with respect to rooted bisimilarity. 
The formalisation brings up and exploits similarities between results
and proofs for different equivalences and preorders. Indeed we have
considered several `unique-solution' results (for various equivalences
and preorders); they share many parts of the proofs, but present a few
delicate and subtle differences in a few points. In a paper-pencil
proof, checking all details would be long and \hl{error-prone},
\hl{especially in cases where the proofs are similar to each other or
  when there are long case analyses to be carried out.}
Some of the textbook proofs are even wrong, e.g. Milner's
proof of the unique-solution theorem for $wb$
(Theorem~\ref{t:Mil89s3}).
For some other textbook proofs, even they are correct, sometimes they actually
do not need all information from the antecedents, which can be further
weakened. This kind of improvements can be easily found during the
formalisation work.
For our CCS formalisation, we believe that all the definitions and
theorem statements are easy to read and
understand, as they are very close to their original statements in
textbooks~\cite{Gorrieri:2015jt,Mil89}.

Formalising other equivalences and preorders could also be considered,
notably the \hl{trace-based} equivalences, as well as more refined process
calculi such as value-passing CCS.
% (e.g.~exploiting the type variable
%of actions).}
%
On another research line, one could examine the formalisation of a different
approach \cite{DurierHS17} to unique
solutions, in which the use of contraction is
replaced by semantic conditions on process transitions such as
divergence. 
%We hope this work also inspires new formalisations on other process calculi.

% Further plan on the formalisation mainly includes: 1) the extension to
% multi-variable equation/contractions. 2) the support of recursion
% operators in 
% CCS context and expressions.

>>>>

In this paper, beside the introduction of rooted contraction and its
unique solution theorems as a theoretical extension
of~\cite{sangiorgi2017equations},
we highlighted a comprehensive formalisation of the theory of CCS in the 
HOL4 theorem prover. In particular, the formalisation supports four
methods for establishing (strong and weak) bisimilarities:
\begin{enumerate}
\item constructing a bisimulation (the standard bisimulation proof
method);
\item constructing a `bisimulation up-to'; 
\item employing algebraic laws;
\item constructing a system of equations or contractions
(i.e., the `unique-solution' method)
\end{enumerate}

The formalisation has actually focused on the theory of
unique solution of equations and contractions, both \univariate and
\multivariate cases. It    
 has allowed us to further develop the theory,
notably the basic properties of rooted contraction, and the unique
solution theorem for it with respect to rooted bisimilarity. 
The formalisation brings up and exploits similarities between results
and proofs for different equivalences and preorders. Indeed we have
considered several `unique-solution' results (for various equivalences
and preorders); they share many parts of the proofs, but present a few
delicate and subtle differences in a few points. In a paper-pencil
proof, checking all details would be long and \hl{error-prone},
\hl{especially in cases where the proofs are similar to each other or
  when there are long case analyses to be carried out.}
\hl{Some of the textbook proofs were even wrong and the correct version is
known but not available in the literature. This is the case of Milner's
proof of the unique-solution theorem for $wb$}
(Theorem~\ref{t:Mil89s3}). \hl{Now we have fully verified formal proofs for
people who can read them.}
For some other textbook proofs, even they are correct, sometimes they actually
do not need all information from the antecedents, which can be further
weakened. This kind of improvements can be easily found during the
formalisation work.
For our CCS formalisation, we believe that all the definitions and
theorem statements are easy to read and
understand, as they are very close to their original statements in
textbooks~\cite{Gorrieri:2015jt,Mil89}.

For the future work, formalising other equivalences and preorders could also be considered,
notably the \hl{trace-based} equivalences, as well as more refined process
calculi such as value-passing CCS.
% (e.g.~exploiting the type variable
%of actions).}
%
On another research line, one could examine the formalisation of a different
approach \cite{DurierHS17} to unique
solutions, in which the use of contraction is
replaced by semantic conditions on process transitions such as
divergence.
%
Finally, the disjointness between free and bound variables in the current
\multivariate formalisation is certainly a limitation that can be
removed in the future. If needed, extending the current work to
infinite number of equations and variables is also feasible.

### 2019-09-15 ic-journal/context.tex
text
Minor fixes, closing the paper
Literally handled ALL review issues. DONE.

We have chosen to use $\lambda$-expressions (with the type
``\HOLinline{\ensuremath{(}\ensuremath{\alpha}, \ensuremath{\beta}\ensuremath{)} \HOLTyOp{CCS} \HOLTokenTransEnd \ensuremath{(}\ensuremath{\alpha}, \ensuremath{\beta}\ensuremath{)} \HOLTyOp{CCS}}'')
to represent \emph{multi-hole contexts}.
%
This choice has a significant advantage over \emph{one-hole
contexts}, as each hole corresponds to one
appearance of the \emph{same} variable in single-variable
expressions (or equations). Thus \emph{contexts} can be directly used in
formulating the unique solution of equations theorems in
single-variable cases. The precise definition is given inductively:
\begin{alltt}
\HOLTokenTurnstile{} \HOLConst{CONTEXT} \ensuremath{(}\HOLTokenLambda{}\HOLBoundVar{t}. \HOLBoundVar{t}\ensuremath{)} \HOLSymConst{\HOLTokenConj{}} \ensuremath{(}\HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{p}. \HOLConst{CONTEXT} \ensuremath{(}\HOLTokenLambda{}\HOLBoundVar{t}. \HOLBoundVar{p}\ensuremath{)}\ensuremath{)} \HOLSymConst{\HOLTokenConj{}}
   \ensuremath{(}\HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{a} \HOLBoundVar{e}. \HOLConst{CONTEXT} \HOLBoundVar{e} \HOLSymConst{\HOLTokenImp{}} \HOLConst{CONTEXT} \ensuremath{(}\HOLTokenLambda{}\HOLBoundVar{t}. \HOLBoundVar{a}\HOLSymConst{\ensuremath{\ldotp}}\HOLBoundVar{e} \HOLBoundVar{t}\ensuremath{)}\ensuremath{)} \HOLSymConst{\HOLTokenConj{}}
   \ensuremath{(}\HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{e\sb{\mathrm{1}}} \HOLBoundVar{e\sb{\mathrm{2}}}. \HOLConst{CONTEXT} \HOLBoundVar{e\sb{\mathrm{1}}} \HOLSymConst{\HOLTokenConj{}} \HOLConst{CONTEXT} \HOLBoundVar{e\sb{\mathrm{2}}} \HOLSymConst{\HOLTokenImp{}} \HOLConst{CONTEXT} \ensuremath{(}\HOLTokenLambda{}\HOLBoundVar{t}. \HOLBoundVar{e\sb{\mathrm{1}}} \HOLBoundVar{t} \HOLSymConst{\ensuremath{+}} \HOLBoundVar{e\sb{\mathrm{2}}} \HOLBoundVar{t}\ensuremath{)}\ensuremath{)} \HOLSymConst{\HOLTokenConj{}}
   \ensuremath{(}\HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{e\sb{\mathrm{1}}} \HOLBoundVar{e\sb{\mathrm{2}}}. \HOLConst{CONTEXT} \HOLBoundVar{e\sb{\mathrm{1}}} \HOLSymConst{\HOLTokenConj{}} \HOLConst{CONTEXT} \HOLBoundVar{e\sb{\mathrm{2}}} \HOLSymConst{\HOLTokenImp{}} \HOLConst{CONTEXT} \ensuremath{(}\HOLTokenLambda{}\HOLBoundVar{t}. \HOLBoundVar{e\sb{\mathrm{1}}} \HOLBoundVar{t} \HOLSymConst{\ensuremath{\mid}} \HOLBoundVar{e\sb{\mathrm{2}}} \HOLBoundVar{t}\ensuremath{)}\ensuremath{)} \HOLSymConst{\HOLTokenConj{}}
   \ensuremath{(}\HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{L} \HOLBoundVar{e}. \HOLConst{CONTEXT} \HOLBoundVar{e} \HOLSymConst{\HOLTokenImp{}} \HOLConst{CONTEXT} \ensuremath{(}\HOLTokenLambda{}\HOLBoundVar{t}. \ensuremath{(\nu}\HOLBoundVar{L}\ensuremath{)} \ensuremath{(}\HOLBoundVar{e} \HOLBoundVar{t}\ensuremath{)}\ensuremath{)}\ensuremath{)} \HOLSymConst{\HOLTokenConj{}}
   \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{rf} \HOLBoundVar{e}. \HOLConst{CONTEXT} \HOLBoundVar{e} \HOLSymConst{\HOLTokenImp{}} \HOLConst{CONTEXT} \ensuremath{(}\HOLTokenLambda{}\HOLBoundVar{t}. \HOLConst{relab} \ensuremath{(}\HOLBoundVar{e} \HOLBoundVar{t}\ensuremath{)} \HOLBoundVar{rf}\ensuremath{)}\hfill{[CONTEXT_rules]}
\end{alltt}

@@

\hl{Many lemmas about above concepts} (\HOLinline{\HOLConst{CONTEXT}}, \HOLinline{\HOLConst{WG}},
\HOLinline{\HOLConst{SEQ}}, etc.) have then to be proven for the
 relationships among these kinds of contexts and properties about their
 compositions. These proofs are usually tedious and long,  due to multiple levels of
 inductions on the structure of the contexts.

@@

Also,
\begin{alltt}
\HOLTokenTurnstile{} \HOLConst{precongruence} \HOLConst{OBS_contracts}\hfill{[OBS_contracts_precongruence]}
\end{alltt}

>>>>

We have chosen to use $\lambda$-expressions (with the type
``\HOLinline{\ensuremath{(}\ensuremath{\alpha}, \ensuremath{\beta}\ensuremath{)} \HOLTyOp{CCS} \HOLTokenTransEnd \ensuremath{(}\ensuremath{\alpha}, \ensuremath{\beta}\ensuremath{)} \HOLTyOp{CCS}}'')
to represent \emph{multi-hole contexts}.
%
This choice has a significant advantage over \emph{one-hole
contexts}, as each hole corresponds to one
appearance of the \emph{same} variable in single-variable
expressions (or equations). Thus \emph{contexts} can be directly used in
formulating the unique solution of equations theorems in
single-variable cases. The precise definition is given inductively:
\begin{alltt}
\HOLTokenTurnstile{} \HOLConst{CONTEXT} \ensuremath{(}\HOLTokenLambda{}\HOLBoundVar{t}. \HOLBoundVar{t}\ensuremath{)} \HOLSymConst{\HOLTokenConj{}} \ensuremath{(}\HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{p}. \HOLConst{CONTEXT} \ensuremath{(}\HOLTokenLambda{}\HOLBoundVar{t}. \HOLBoundVar{p}\ensuremath{)}\ensuremath{)} \HOLSymConst{\HOLTokenConj{}}
   \ensuremath{(}\HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{a} \HOLBoundVar{e}. \HOLConst{CONTEXT} \HOLBoundVar{e} \HOLSymConst{\HOLTokenImp{}} \HOLConst{CONTEXT} \ensuremath{(}\HOLTokenLambda{}\HOLBoundVar{t}. \HOLBoundVar{a}\HOLSymConst{\ensuremath{\ldotp}}\HOLBoundVar{e} \HOLBoundVar{t}\ensuremath{)}\ensuremath{)} \HOLSymConst{\HOLTokenConj{}}
   \ensuremath{(}\HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{e\sb{\mathrm{1}}} \HOLBoundVar{e\sb{\mathrm{2}}}. \HOLConst{CONTEXT} \HOLBoundVar{e\sb{\mathrm{1}}} \HOLSymConst{\HOLTokenConj{}} \HOLConst{CONTEXT} \HOLBoundVar{e\sb{\mathrm{2}}} \HOLSymConst{\HOLTokenImp{}} \HOLConst{CONTEXT} \ensuremath{(}\HOLTokenLambda{}\HOLBoundVar{t}. \HOLBoundVar{e\sb{\mathrm{1}}} \HOLBoundVar{t} \HOLSymConst{\ensuremath{+}} \HOLBoundVar{e\sb{\mathrm{2}}} \HOLBoundVar{t}\ensuremath{)}\ensuremath{)} \HOLSymConst{\HOLTokenConj{}}
   \ensuremath{(}\HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{e\sb{\mathrm{1}}} \HOLBoundVar{e\sb{\mathrm{2}}}. \HOLConst{CONTEXT} \HOLBoundVar{e\sb{\mathrm{1}}} \HOLSymConst{\HOLTokenConj{}} \HOLConst{CONTEXT} \HOLBoundVar{e\sb{\mathrm{2}}} \HOLSymConst{\HOLTokenImp{}} \HOLConst{CONTEXT} \ensuremath{(}\HOLTokenLambda{}\HOLBoundVar{t}. \HOLBoundVar{e\sb{\mathrm{1}}} \HOLBoundVar{t} \HOLSymConst{\ensuremath{\mid}} \HOLBoundVar{e\sb{\mathrm{2}}} \HOLBoundVar{t}\ensuremath{)}\ensuremath{)} \HOLSymConst{\HOLTokenConj{}}
   \ensuremath{(}\HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{L} \HOLBoundVar{e}. \HOLConst{CONTEXT} \HOLBoundVar{e} \HOLSymConst{\HOLTokenImp{}} \HOLConst{CONTEXT} \ensuremath{(}\HOLTokenLambda{}\HOLBoundVar{t}. \ensuremath{(\nu}\HOLBoundVar{L}\ensuremath{)} \ensuremath{(}\HOLBoundVar{e} \HOLBoundVar{t}\ensuremath{)}\ensuremath{)}\ensuremath{)} \HOLSymConst{\HOLTokenConj{}}
   \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{rf} \HOLBoundVar{e}. \HOLConst{CONTEXT} \HOLBoundVar{e} \HOLSymConst{\HOLTokenImp{}} \HOLConst{CONTEXT} \ensuremath{(}\HOLTokenLambda{}\HOLBoundVar{t}. \HOLConst{relab} \ensuremath{(}\HOLBoundVar{e} \HOLBoundVar{t}\ensuremath{)} \HOLBoundVar{rf}\ensuremath{)}\hfill{[CONTEXT_rules]}
\end{alltt}
Notice in above inductive definitions, for any CCS term $p$,
\HOLinline{\HOLTokenLambda{}\HOLBoundVar{t}.\\\;\HOLFreeVar{p}} is a valid context. In this case there is no
hole in the context. On the other hand, if any CCS term with recursion
could be a context, the recursion must be a subterm of this $p$.

@@

\hlD{Many lemmas about the  above concepts} (\HOLinline{\HOLConst{CONTEXT}}, \HOLinline{\HOLConst{WG}},
\HOLinline{\HOLConst{SEQ}}, etc.) have then to be proven for the
 relationships among these kinds of contexts and properties about their
 compositions. These proofs are usually tedious and long,  due to multiple levels of
 inductions on the structure of the contexts.

@@

Rooted contraction, on the other hand, is a precongruence:
\begin{alltt}
\HOLTokenTurnstile{} \HOLConst{precongruence} \HOLConst{OBS_contracts}\hfill{[OBS_contracts_precongruence]}
\end{alltt}
